---
title: "Statistics Research Project: Tennis Point Streaks"
author: "Muhammed Altindal"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
rm(list = ls())
library(dplyr)
library(ggplot2)
```

# Introduction

The goal of this project is to study short-term momentum in tennis by examining sequences of consecutive points won by the same player. Using point-by-point data, I focus on identifying and describing winning streaks within matches. The main purpose is descriptive: to see how long these streaks tend to be, whether long streaks are rare, and how streak patterns vary across matches and players.

# Loading the Data

```{r}
matches <- read.csv("2011-frenchopen-matches.csv", stringsAsFactors = FALSE)
points  <- read.csv("2011-frenchopen-points.csv", stringsAsFactors = FALSE)
```

# Initial Inspection

Before constructing point-winning streaks, I first inspect the structure of both datasets.

```{r}
dim(matches)
dim(points)

names(matches)
names(points)

str(matches)
str(points)
```

# Data Preparation

To analyze point-winning streaks, I attached player names to each point by joining the points dataset with the matches dataset using `match_id`. Then I create readable player-name variables for both the point winner and the point server.

```{r}
matches_small <- matches %>%
  select(match_id, player1, player2) %>%
  rename(
    P1_name = player1,
    P2_name = player2
  )

points_named <- points %>%
  left_join(matches_small, by = "match_id") %>%
  mutate(
    PointWinnerName = case_when(
      PointWinner == 1 ~ P1_name,
      PointWinner == 2 ~ P2_name,
      TRUE ~ NA_character_
    ),
    PointServerName = case_when(
      PointServer == 1 ~ P1_name,
      PointServer == 2 ~ P2_name,
      TRUE ~ NA_character_
    )
  )

```

# The cleaning step

In this step, I cleaned the point-level dataset by keeping only valid observations where a point was actually played. I remove rows with invalid winner codes and exclude non-point records (such as match-start rows) by filtering out zero or missing point numbers. Finally, I sorted the data within each match to ensure that points are in the correct chronological order for streak analysis.

```{r}
points_clean <- points_named %>%
  filter(PointWinner %in% c(1, 2)) %>%      # keep only actual point winners
  filter(!is.na(PointNumber)) %>%           # remove missing point numbers
  filter(PointNumber > 0) %>%               # remove match-start / non-point rows
  filter(!is.na(match_id)) %>%
  arrange(match_id, SetNo, GameNo, PointNumber)
```

#Arranging winning player column

```{r}
points_clean <- points_named %>%
  filter(PointWinner %in% c(1, 2)) %>%
  filter(!is.na(PointNumber)) %>%
  filter(PointNumber > 0) %>%
  filter(!is.na(match_id)) %>%
  mutate(
    PointWinnerName = case_when(
      PointWinner == 1 ~ P1_name,
      PointWinner == 2 ~ P2_name
    ),
    PointServerName = case_when(
      PointServer == 1 ~ P1_name,
      PointServer == 2 ~ P2_name
    )
  ) %>%
  arrange(match_id, SetNo, GameNo, PointNumber)

points_clean <- points_clean %>%
  relocate(PointWinnerName, .after = PointWinner) %>%
  relocate(PointServerName, .after = PointServer)

#points_clean
```

# Data Quality Checks

Before constructing streaks, I check whether the merge worked properly, whether the winner/server codes look valid, whether any point records are duplicated, and whether the cleaned dataset contains missing player names.

#Is NA? 
```{r}
sum(is.na(points_named$P1_name))
sum(is.na(points_named$P2_name))
```


```{r}
table(points$PointWinner, useNA = "ifany")
table(points$PointServer, useNA = "ifany")
```
The raw point data mainly uses `1` and `2` as expected, but it also contains some `0` values and a small number of missing values. This suggests that the dataset includes a few non-point or incomplete rows, which were removed during data cleaning.

```{r}
sum(is.na(points_clean$PointWinnerName))
sum(is.na(points_clean$PointServerName))
```

#Duplicate check

```{r}
duplicate_points_raw <- points %>%
  count(match_id, SetNo, GameNo, PointNumber) %>%
  filter(n > 1)

duplicate_points_raw
nrow(duplicate_points_raw)
```
#Inspecting point ordering

```{r}
example_match <- points_clean$match_id[1]

points_clean %>%
  filter(match_id == example_match) %>%
  select(match_id, SetNo, GameNo, PointNumber, PointWinner, PointWinnerName) %>%
  head(30)
```


```{r}
n_total   <- nrow(points_named)
n_clean   <- nrow(points_clean)
n_removed <- n_total - n_clean
pct_removed <- round(100 * n_removed / n_total, 2)

n_total
n_clean
n_removed
pct_removed
```
Only a very small proportion of observations (approximately 0.57%) were removed during cleaning, indicating that the raw dataset was largely complete and consistent.

# Constructing Point-Winning Streaks

To identify streaks, I group the cleaned data by match and then mark a new streak whenever the point winner changes from one point to the next. Consecutive points won by the same player are assigned the same streak ID.

```{r}
streaks <- points_clean %>%
  group_by(match_id) %>%
  mutate(
    new_streak = ifelse(
      row_number() == 1, 1,
      ifelse(PointWinner != lag(PointWinner), 1, 0)
    ),
    streak_id = cumsum(new_streak)
  ) %>%
  group_by(match_id, streak_id) %>%
  summarise(
    winner_code   = first(PointWinner),
    winner_name   = first(PointWinnerName),
    P1_name       = first(P1_name),
    P2_name       = first(P2_name),
    streak_length = n(),
    .groups = "drop"
  ) %>%
  mutate(
    loser_name = ifelse(winner_name == P1_name, P2_name, P1_name)
  )
```

#After constructing the streaks, I check whether the results look reasonable.

```{r}
summary(streaks$streak_length)
table(streaks$streak_length)

sum(is.na(streaks$winner_name))
sum(is.na(streaks$loser_name))

streaks %>%
  filter(streak_length <= 0)
streaks
```
Streak lengths should always be positive. Missing player names or zero-length streaks would indicate a problem in the construction process.

#The following output summarizes the distribution of streak lengths.

```{r}
streaks %>%
  summarise(
    Number_of_streaks = n(),
    Mean_streak_length = mean(streak_length),
    Median_streak_length = median(streak_length),
    Max_streak_length = max(streak_length)
  )
```

#To visualize the distribution of streak lengths, I begin with a histogram.

```{r}
ggplot(streaks, aes(x = streak_length)) +
  geom_histogram(binwidth = 1, boundary = 0, closed = "left") +
  labs(
    title = "Distribution of Point-Winning Streak Lengths",
    x = "Streak Length",
    y = "Frequency"
  ) +
  theme_minimal()
```
## Added a boxplot

```{r}
ggplot(streaks, aes(y = streak_length)) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Point-Winning Streak Lengths",
    y = "Streak Length"
  ) +
  theme_minimal()
```

# Inspecting Long Streaks

Long streaks may be especially interesting because they could reflect strong short-term momentum or unusual game situations.

```{r}
long_streaks <- streaks %>%
  filter(streak_length >= 6) %>%
  arrange(desc(streak_length)) %>%
  relocate(streak_length, .after = winner_name)

long_streaks
```

Here, I identify and examine long point-winning streaks to explore whether unusually long sequences may indicate short-term momentum.

Tennis point outcomes show some clustering (streaks), but most streaks are short, suggesting limited evidence of strong momentum effects.



```{r}
saveRDS(streaks, "streaks.rds")
```

# PART 2

Creating graph, looking for longest streak match. 

```{r}
sharapova_longest <- streaks %>%
  filter(winner_name == "Maria Sharapova") %>%
  arrange(desc(streak_length))

sharapova_longest
```

```{r}
sharapova_top_match <- sharapova_longest %>%
  slice(1)

sharapova_top_match
```
```{r}
selected_match_id <- "2011-frenchopen-2149"

match_plot_data <- points_clean %>%
  filter(match_id == selected_match_id) %>%
  mutate(
    point_index = row_number(),
    winner_y = ifelse(PointWinner == 1, 1, 0),
    server_y = ifelse(PointServer == 1, 1, 0),
    server_won = PointWinner == PointServer,
    streak_break = ifelse(row_number() == 1, FALSE, PointWinner != lag(PointWinner))
  ) %>%
  mutate(
    one_point_interruption = ifelse(
      row_number() %in% c(1, n()), FALSE,
      lag(PointWinner) == lead(PointWinner) & PointWinner != lag(PointWinner)
    )
  )
```

```{r}
match_plot_data %>%
  select(P1_name, P2_name) %>%
  distinct()
```
```{r}
library(dplyr)
library(ggplot2)

# 1) Find the match in which Maria Sharapova has her longest streak
sharapova_longest <- streaks %>%
  filter(winner_name == "Maria Sharapova") %>%
  arrange(desc(streak_length))

sharapova_top_match <- sharapova_longest %>%
  slice(1)

selected_match_id <- sharapova_top_match$match_id[1]

# 2) Prepare point-level data for the selected match
match_plot_data <- points_clean %>%
  filter(match_id == selected_match_id) %>%
  arrange(SetNo, GameNo, PointNumber) %>%
  mutate(
    point_index = row_number(),
    winner_y = ifelse(PointWinner == 1, 1.00, 0.94),
    point_score = paste0(P1Score, "-", P2Score),
    streak_break = ifelse(row_number() == 1, FALSE, PointWinner != lag(PointWinner))
  ) %>%
  mutate(
    one_point_interruption = ifelse(
      row_number() %in% c(1, n()),
      FALSE,
      lag(PointWinner) == lead(PointWinner) & PointWinner != lag(PointWinner)
    ),
    next_x = lead(point_index),
    next_y = lead(winner_y)
  )

# 3) Extract player labels
p1_label <- unique(match_plot_data$P1_name)
p2_label <- unique(match_plot_data$P2_name)

# 4) Create game-level summary for serve blocks and game scores
game_summary <- match_plot_data %>%
  group_by(SetNo, GameNo) %>%
  summarise(
    game_start = min(point_index),
    game_end   = max(point_index),
    game_mid   = (game_start + game_end) / 2,
    server_name = first(PointServerName),
    p1_games = max(P1GamesWon),
    p2_games = max(P2GamesWon),
    game_score = paste0(p1_games, "-", p2_games),
    .groups = "drop"
  ) %>%
  mutate(
    serve_label = ifelse(server_name == p1_label, "P1 serve", "P2 serve")
  )

# 5) Create set-level summary for final set scores
set_summary <- match_plot_data %>%
  group_by(SetNo) %>%
  summarise(
    set_start = min(point_index),
    set_end   = max(point_index),
    set_mid   = (set_start + set_end) / 2,
    p1_games = max(P1GamesWon),
    p2_games = max(P2GamesWon),
    n_games = n_distinct(GameNo),
    set_score = paste0(p1_games, "-", p2_games),
    .groups = "drop"
  ) %>%
  mutate(
    set_label = paste0(
      "Set ", SetNo,
      " (Games 1-", n_games, "): ",
      set_score
    )
  )

# 6) Add an initial 0-0 score label
initial_score <- data.frame(
  x = min(match_plot_data$point_index),
  y = 0.905,
  label = "0-0"
)

# 7) Build the plot
ggplot() +
  
  # Serve blocks by game
  geom_rect(
    data = game_summary,
    aes(
      xmin = game_start - 0.5,
      xmax = game_end + 0.5,
      ymin = 0.925,
      ymax = 1.015,
      fill = server_name
    ),
    alpha = 0.12
  ) +
  
  # Segments connecting each point to the next point
  geom_segment(
    data = match_plot_data %>% filter(!is.na(next_x)),
    aes(
      x = point_index,
      xend = next_x,
      y = winner_y,
      yend = next_y,
      color = PointWinnerName
    ),
    linewidth = 1
  ) +
  
  # Winner points
  geom_point(
    data = match_plot_data,
    aes(x = point_index, y = winner_y, color = PointWinnerName),
    size = 2.2
  ) +
  
  # Mark streak breaks with X
  geom_point(
    data = subset(match_plot_data, streak_break),
    aes(x = point_index, y = winner_y),
    shape = 4,
    size = 3,
    stroke = 1.1,
    color = "black"
  ) +
  
  # Mark one-point interruptions with yellow points
  geom_point(
    data = subset(match_plot_data, one_point_interruption),
    aes(x = point_index, y = winner_y),
    shape = 21,
    size = 3.2,
    stroke = 1,
    fill = "yellow",
    color = "black"
  ) +
  
  # Point-by-point tennis score labels
  geom_text(
    data = match_plot_data,
    aes(x = point_index, y = 0.915, label = point_score),
    size = 2.2,
    angle = 90,
    vjust = 0.5,
    color = "gray25"
  ) +
  
  # Dashed vertical lines for game boundaries
  geom_vline(
    data = game_summary,
    aes(xintercept = game_end + 0.5),
    linetype = "dashed",
    color = "gray50",
    linewidth = 0.4
  ) +
  
  # Game score labels
  geom_text(
    data = game_summary,
    aes(x = game_mid, y = 0.905, label = game_score),
    size = 3.2,
    fontface = "bold"
  ) +
  
  # Initial 0-0 score
  geom_text(
    data = initial_score,
    aes(x = x - 1, y = y, label = label),
    size = 3.2,
    fontface = "bold"
  ) +
  
  # Serve labels at the top of each game
  geom_text(
    data = game_summary,
    aes(x = game_mid, y = 1.02, label = serve_label),
    size = 3
  ) +
  
  # Solid vertical lines for set boundaries
  geom_vline(
    data = set_summary,
    aes(xintercept = set_end + 0.5),
    linetype = "solid",
    color = "black",
    linewidth = 0.6
  ) +
  
  # Set labels at the very top
  geom_text(
    data = set_summary,
    aes(
      x = set_mid,
      y = 1.045,
      label = set_label
    ),
    size = 4,
    fontface = "bold"
  ) +
  
  scale_y_continuous(
    breaks = c(0.94, 1.00),
    labels = c(p2_label, p1_label),
    limits = c(0.90, 1.055)
  ) +
  
  labs(
    title = paste("Point Sequence with Serve Context:", selected_match_id),
    x = NULL,
    y = NULL,
    color = "Point Winner",
    fill = "Server",
    caption = "X = Streak break (change in point winner)   |   Yellow point = One-point interruption (temporary break in streak)"
  ) +
  
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.caption = element_text(size = 10, hjust = 0)
  )
```
#Hypothesis thinking part

```{r}
observed_match <- match_plot_data %>%
  select(point_index, PointServer, PointWinner)

observed_streaks <- observed_match %>%
  mutate(
    new_streak = ifelse(
      row_number() == 1,
      1,
      ifelse(PointWinner != lag(PointWinner), 1, 0)
    ),
    streak_id = cumsum(new_streak)
  ) %>%
  group_by(streak_id) %>%
  summarise(
    streak_length = n(),
    .groups = "drop"
  )

observed_max_streak <- max(observed_streaks$streak_length)
observed_n_long6 <- sum(observed_streaks$streak_length >= 6)

observed_max_streak
observed_n_long6
```

#guessing/finding serve advantage

```{r}
p_server_match <- mean(match_plot_data$PointWinner == match_plot_data$PointServer)
p_server_match
```
#simulated match
```{r}
simulate_match_winners <- function(server_vector, p_server) {
  simulated_winner <- sapply(server_vector, function(s) {
    server_wins <- rbinom(1, 1, p_server)
    if (server_wins == 1) {
      return(s)
    } else {
      return(ifelse(s == 1, 2, 1))
    }
  })
  
  return(simulated_winner)
}
```

#smilated streak statistics
```{r}
get_streak_stats <- function(winner_vector) {
  streak_df <- data.frame(PointWinner = winner_vector) %>%
    mutate(
      new_streak = ifelse(
        row_number() == 1,
        1,
        ifelse(PointWinner != lag(PointWinner), 1, 0)
      ),
      streak_id = cumsum(new_streak)
    ) %>%
    group_by(streak_id) %>%
    summarise(
      streak_length = n(),
      .groups = "drop"
    )
  
  c(
    max_streak = max(streak_df$streak_length),
    n_long6 = sum(streak_df$streak_length >= 6),
    mean_streak = mean(streak_df$streak_length)
  )
}
```

#Runninf for 1000 times 
```{r}
set.seed(123)

server_vector <- observed_match$PointServer
n_sim <- 1000

sim_matrix <- replicate(n_sim, {
  sim_winners <- simulate_match_winners(server_vector, p_server_match)
  get_streak_stats(sim_winners)
})

sim_results <- as.data.frame(t(sim_matrix))
#sim_results
```
#p-value of simulations
```{r}
p_value_max <- mean(sim_results$max_streak >= observed_max_streak)
p_value_long6 <- mean(sim_results$n_long6 >= observed_n_long6)

p_value_max
p_value_long6
```
The observed maximum streak is significantly larger than expected under a serve-adjusted chance model.
Long streak counts are somewhat unusual, but not strongly significant at the 5% level

The observed match contains an unusually long maximum streak that is unlikely under a serve-adjusted chance model. However, the total number of long streaks (length ≥ 6) is not strongly significant. This suggests that the match includes one especially extreme run, rather than a broad overall excess of long streaks.

